Formalization of Network Flow Algorithms

### 2017-02-28 paper/root.tex
Iteration #2 over paper

\usepackage{etex}
\pagestyle{plain} % turn on page numbers
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
% \usepackage{mathpartir}
%\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{multirow}
\usepackage{pgfplots}

@@

\input{lstisabelle}
\lstset{basicstyle=\footnotesize\ttfamily\slshape}
\lstset{captionpos=b}
\lstset{numberbychapter=false}

@@

\institute{
  Peter Lammich \at
  Institut f\"ur Informatik,
  Technische Universit\"at M\"unchen \\
  \email{lammich@in.tum.de} %
\and
  S. Reza Sefidgar \at
  Institute of Information Security,
  Department of Computer Science,
  ETH Zurich, Switzerland \\
  \email{reza.sefidgar@inf.ethz.ch} %
}

@@

\begin{abstract}
We present a formalization of two classical algorithms for computing the maximum flow in a network:
The Edmonds-Karp algorithm and Goldberg's Generic Push-Relabel algorithm. We prove the correctness and complexity of the algorithms.
Our formal proof closely follows a standard textbook proof, and is accessible even without being
an expert in Isabelle/HOL --- the interactive theorem prover used for the formalization.

Using stepwise refinement, we obtain the Fifo-Push-Relabel algorithm with gap heuristics. 
Further refinement yields verified efficient implementations of both, Edmonds-Karp and Fifo-Push-Relabel, 
which compare well to unverified reference implementations in Java.

\keywords{Maxflow Algorithm \and Edmonds-Karp Algorithm \and Push-Relabel Algorithm \and Formal Verification \and Isabelle/HOL \and Stepwise Refinement}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
The basic idea is to repeatedly search for \emph{augmenting path} along which more flow can be moved from the source to the sink.
An important instance is the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time ($O(VE^2)$) for the general case of networks with real-valued capacities.
Dinitz~\cite{Di06} generalized the idea of augmenting paths to blocking flows, obtaining an $O(V^2E)$ algorithm. 
Karzanov~\cite{Ka74} was the first to propose an $O(V^3)$ algorithm based on the idea of \emph{preflows}. Based on Karzanov's ideas, Goldberg et al.\ proposed
the generic push-relabel algorithm, which consists of repeated application of two local operations, push and relabel, in any order.
Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
specific variants of the algorithms achieve even lower complexities up to $O(V^2\sqrt{E})$.

In this paper, we present a formal verification of the Edmonds-Karp algorithm and of the generic push-relabel algorithm.
We prove correctness of the algorithms as well as the time complexity bounds of $O(VE^2)$ and $O(V^2E)$, respectively. 
The formalization is conducted in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification to separate the abstract algorithmic ideas from
the implementation details: After proving the min-cut-max-flow theorem, it is straightforward to verify a generic version of the Ford-Fulkerson algorithm,
which we then refine to Edmonds-Karp algorithm and finally to an efficient implementation. Similar, the generic push-relabel algorithm is refined to 
the fifo-push-relabel algorithm, which, in turn, is refined to an efficient implementation.
The abstract parts of our verification closely follow the textbook presentation of Cormen et al.~\cite{CLRS09}. 
Using the Isar~\cite{Wenzel99} proof language, we were able to produce proofs that are accessible even to non-Isabelle experts.

This paper is an extended version of our conference paper~\cite{LaSe16}, which only covered the Edmonds-Karp algorithm.

While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial time maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation. Moreover, this paper is a case study on elegantly formalizing algorithms.

The rest of this paper is structured as follows: In Section~\ref{sec:background} we give a short informal introduction to the Ford-Fulkerson method.
In Section~\ref{sec:abs-formalization}, we report on our formalization of the abstract method. 
Section~\ref{sec:refinement} gives a brief overview of the Isabelle Refinement Framework~\cite{LaTu12,La12}, which supports stepwise refinement based algorithm development in Isabelle/HOL. In Section~\ref{sec:edka}, we report on our instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm and the proof of its complexity.
Section~\ref{sec:executable} reports on the further refinement steps required to yield an efficient implementation. Section~\ref{sec:benchmark} reports on 
benchmarking our implementation against a reference implementation of the Edmonds-Karp algorithm from Sedgewick et al.~\cite{SeWa11}. 
Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work. The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}.

@@

A \emph{preflow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Non-generation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
Intuitively, a \emph{preflow} must not send more flow over an edge than its capacity allows, and only the source node may generate flow. 
The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all vertices except $s$ and $t$, the excess flow is zero.
Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

@@

The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is usually stated as the following three statements being equivalent:
\begin{enumerate}
\item $f$ is a maximum flow in a network $G$.
\item there is no augmenting path in the residual graph $G_f$.
\item there is a cut $C$ in $G$ such that the capacity of $C$ is equal to the value of $f$.
\end{enumerate}

@@

In a typical program development, one first comes up with an initial version \isai{m_0} of
the algorithm and its specification \isai{P,Q}, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions \isai{m$_i$} of the algorithm,
proving \isai{m$_i$ <= \<Down>R$_i$ m$_{i-1}$}. Using transitivity and composability of data refinement, one 
gets \isai{P ==> m$_i$ <= \<Down>R$_i$...R$_1$ spec Q}, showing the correctness of the refined algorithm.
If no data refinement is performed, \isai{R$_i$} is set to the identity relation, in which case \isai{\<Down>R$_i$} becomes the identity function.

@@

\section{The Generic Push-Relabel Algorithm}  
  We present our formalization of the generic push relabel algorithm of Goldberg et al.~\cite{GoTa88}. 
  Again, our formalization closely follows the textbook presentation of Cormen et al.~\cite{CLRS09}.
  The push-relabel algorithm maintains a preflow $f$ and a height-labeling $l: V \to \mathbb {N}$ of the nodes, 
  with the invariant that for every edge $(u, v)$ of the residual graph we have that $l(u) \le l(v) + 1$. 
  Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
  By generalizing the labeling invariant to paths, we get that for each path $p$ from $u$ to $v$, we have $l(u) \le l(v) + |p|$.
  This implies that there cannot be a path from $s$ to $t$: The length of any (simple) path is less than $|V|$, however, a valid labeling 
  implies a length greater than or equal to $|V|$.

  A node that has positive excess is called \emph{active}. Upon termination of the algorithm, there will be no active nodes. 
  Thus, the preflow is actually a flow, and as we have no $s$--$t$ paths, the min-cut-max-flow theorem implies that it is maximal.

  Initially, the height of all nodes but the source is $0$, and the outgoing edges of the source get the maximum possible flow, 
  and all other edges get no flow at all. Then, the algorithm repeats the following operations:
  \begin{description}
    \item[push] Send as much flow as possible over an edge $(u,v)$ with $l(u) = l(v)+1$. The maximum possible flow is constrained
      by the residual capacity of the edge, and the excess flow available at $u$.
    \item[relabel] Increase the height of an active (non-sink) node from which no push operation is possible. 
      The height is increased to the minimum value such that a push becomes possible.
  \end{description}
  The algorithm is generic in the sense that it leaves open the order in which the push and relabel operations on the different nodes and edges are performed.
  The algorithm terminates if no more operations apply. Then, there are no active nodes left, and the algorithm has constructed a maximum flow.

\subsection{Formalization of the Generic Algorithm}
For the complexity reasoning, we will count the different types of operations. Thus, it is convenient to formalize the algorithm as a labeled transition system
over flows and heights, the labels indicating the performed operation:
\begin{lstlisting}
  inductive_set pr_lts :: "((preflow\<times>labeling) \<times> op_ty \<times> (preflow\<times>labeling)) set" 
  where
    "push_pre f l e ==> ((f,l),PUSH,(push_eff f e,l)) \<in> pr_lts"
  | "relabel_pre f l u ==> ((f,l),RELABEL,(f,relabel_eff f l u)) \<in> pr_lts"
\end{lstlisting}  
Here, the \isai{inductive_set} construction defines the least relation that can be derived by the specified rules. 
A push operation can be implemented in time $O(1)$, (only the flow on a single edge has to be modified) and a relabel operation can be implemented in time $O(V)$.
(to find the new height, we have to consider the heights of all adjacent nodes in the residual graph) To this end, we define 
\isai{cost PUSH == 1 | cost RELABEL == card(V)}. We prove the following theorem:
\begin{lstlisting}
theorem (in Network) generic_preflow_push_OV2E_and_correct:
  assumes "((f_0, l_0), p, (f, l)) \<in> pr_lts^*" 
  shows "(\<sum>x<-p. cost x) \<le> 38 * (card V)$^2$ * card E"
    and "(f,l)\<notin>Domain pr_lts --> isMaxFlow f"
\end{lstlisting}  
where \isai$(f_0,l_0)$ is the initial flow and height-labeling. Intuitively, the cost of any (partial) run \isai$p$ is bounded by $O(V^2E)$, and if the 
run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), the resulting flow is maximal.

For the proof of this theorem, we follow the presentation of Cormen~\cite{CLRS09}. 
However, at several places, we trade precision of the constant for simplicity of the proof. 
Thus, our constant of $38$ is somewhat higher than what would result from Cormen's proof.

For the correctness argument, we show that the operations of the algorithm maintain a valid labeling, and that existence of an active node implies applicability of 
a push or relabel operation.
The complexity argument is more tricky: First, we show that the labels of each node remain smaller than $2|V|$. This bounds the number of relabel operations to $O(V^2)$.
For the push operations, we first define a push to be \emph{saturating}, if the excess at the source node of the push is not less than the capacity of the edge.
In this case, the edge is completely saturated by the push and disappears from the residual graph. However, the source node of the push remains active.
On the other hand, a non-saturating push deactivates the source node of the push.

We count the number of saturating and non-saturating pushes separately. For saturating pushes, we regard the number of saturating pushes for each edge $(u,v)$ during execution of the algorithm: After a saturating push, the edge has disappeared from the residual graph. Thus, before the next saturating push on the same edge, 
there must be an operation that restores the edge. This can only be a push on the reverse edge $(v,e)$. 
However, this push can only occur after the labeling of $v$ has increased. This can only happen $O(V)$ times, which yields a bound of $O(VE)$ on the number 
of saturating push operations. Finally, the number of non-saturating pushes is bounded by defining a potential function that is decreased by non-saturating pushes,
and increased by at most $O(V)$ by the other operations. The initial potential is $O(V)$, and with the already established bounds on the other operations we get a bound of
$O(V^3 + V^2E) = O(V^2E)$ on the number of non-saturating push operations. Combining the bounds for the different operation types with the costs for the operations 
then yields the desired time complexity bound of $O(V^2E)$.

The simple structure of the generic algorithm allowed us to conveniently define it as labeled transition system, such that the above intuitive arguments could be formalized
as counting different operation types on execution paths of the algorithm. Moreover, the powerful Isabelle/HOL standard library, in particular the lemmas for summations
over finite sets, were very helpful. Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few men weeks to complete.

\section{Implementing the Push-Relabel Algorithm}
The generic algorithm does not specify the order in which the operations should be applied. 
Moreover, choosing the right ordering may further decrease the worst-case time complexity, and, more important, the practical efficiency of the algorithm.

We refine the generic push-relabel algorithm to different implementations. However, these have a more complicated algorithmic structure that
prevents a convenient specification as labeled transition system, and thus a complexity proof in the style of the generic algorithm.
Thus, we specify the algorithms using the Monadic Refinement Framework, and only prove correctness and termination 
(which we get for free, as the complexity bound for the generic algorithm is also valid for all implementations).
We leave it to future work to integrate complexity reasoning tools into the Refinement Framework and prove tighter complexity bounds for the specific implementations.

\subsection{Relabel-to-Front Algorithm}
A general idea of the implementation is to \emph{discharge} a node by repeatedly applying push and relabel operations to the same node, until it gets inactive.
Cormen et al.\ present the relabel to front algorithm, which scans through a list of all nodes, discharging every node as it is visited.
If the node gets relabeled during discharging, it is moved to the front of the list, and the scan of the list is restarted at the beginning of the list.
The algorithm terminates if it reaches the end of the list. 
During discharging a node $u$, the algorithm iterates over the list of all adjacent nodes $v$ in the network, and checks whether there is an edge $(u,v)$ in 
the residual graph. When the discharge operation stops without relabeling $u$, the current position in the list of adjacent nodes is stored, and 
the next discharge operations starts at exactly this position.

@@

Following the outline by Cormen et al., and using the Isabelle Refinement Framework to conveniently describe the abstract algorithm,
the correctness proof offered no surprises. It is about 700 lines of Isabelle text plus 150 lines of general lemmas on topological orderings, 
and took only a few days to develop.

\subsection{FIFO Push-Relabel Algorithm}
While the relabel-to-front algorithm has a theoretical worst case complexity of $O(V^3)$, it seems to be rather inefficient in practice.\footnote{We have not found any implementation based on relabel-to-front, and our own experiments indicate a rather poor performance.}
An alternative implementation is the FIFO push-relabel algorithm, which maintains a first-in-first-out queue of the active nodes. 
In each iteration, it pulls a node from the FIFO, and repeatedly pushes flow out of it.
If no push operations are possible any more, but the node is still active, it is relabeled and appended to the FIFO.

The FIFO push relabel algorithm has several advantages: It's correctness is straightforward to prove, 
the only required additional invariant is that the queue contains exactly the active nodes. 
Moreover, it performs well in practice, and we have found several implementations of it~\cite{ChGo97,ICPC-notebook}.

Our algorithm follows the implementation that we found in the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, 
which uses an additional gap heuristics: If there is a value $0<k<|V|$ such that there is no node that has height $k$, 
all nodes with $k<l(v)<|V|$ may be relabeled to $|V|+1$, preserving a valid labeling. In practice, this may drastically accelerate convergence of the algorithm,
as it prevents unnecessary pushes of flow towards the sink when the path to the sink is already blocked.
The correctness proof of this heuristics is straightforward from the definitions, and also the overall correctness proof 
is straightforward: It's size is about 400 lines of code, and it took roughly a day to complete.

>>>>

\usepackage{etex}
\pagestyle{plain} % turn on page numbers
\usepackage[utf8]{inputenc}
\usepackage{microtype} % Better typesetting for PDFs -- is enabling this ok?
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{eufrak} %The eufrak package is redundant if the amsfonts package is used
% \usepackage{mathpartir}
%\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\usepackage[boxed]{algorithm}
\usepackage{enumerate}
\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{multirow}
\usepackage{pgfplots}

@@

\input{lstisabelle}
\lstset{basicstyle=\footnotesize\ttfamily\slshape}
\lstset{captionpos=b}
\lstset{numberbychapter=false}
\lstset{autogobble=true}

@@

\institute{
  Peter Lammich \at
  Institut f\"ur Informatik,
  Technische Universit\"at M\"unchen, Germany \\
  \email{lammich@in.tum.de} %
\and
  S. Reza Sefidgar \at
  Institute of Information Security,
  Department of Computer Science,
  ETH Zurich, Switzerland \\
  \email{reza.sefidgar@inf.ethz.ch} %
}

@@

\begin{abstract}
We present a formalization of classical algorithms for computing the maximum flow in a network:
The Edmonds-Karp algorithm and the push-relabel algorithm. 
We prove correctness and time complexity of these algorithms.
Our formal proof closely follows a standard textbook proof, and is accessible even without being
an expert in Isabelle/HOL --- the interactive theorem prover used for the formalization.

Using stepwise refinement techniques, we instantiate the generic Ford-Fulkerson method to Edmonds-Karp algorithm, 
and the generic push-relabel algorithm of Goldberg and Tarjan to both, the relabel-to-front and the FIFO push-relabel algorithm.
Further refinement then yields verified efficient implementations of the algorithms, which compare well to unverified reference implementations.

\keywords{Maximum Flow Problem \and Edmonds-Karp Algorithm \and Push-Relabel Algorithm \and Formal Verification \and Isabelle/HOL \and Stepwise Refinement}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
The basic idea is to repeatedly search for \emph{augmenting path} along which more flow can be moved from the source to the sink.
An important instance is the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time ($O(VE^2)$) for the general case of networks with real-valued capacities.
Dinitz~\cite{Di06} generalized the idea of augmenting paths to blocking flows, obtaining an $O(V^2E)$ algorithm. 
Karzanov~\cite{Ka74} was the first to propose an $O(V^3)$ algorithm based on the idea of \emph{preflows}. Based on Karzanov's ideas, Goldberg and Tarjan proposed
the generic push-relabel algorithm~\cite{GoTa88}.
Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
specific variants of the algorithms achieve even lower complexities up to $O(V^2\sqrt{E})$.

In this paper, we present a formal verification of the Edmonds-Karp algorithm and of the generic push-relabel algorithm.
We prove correctness of the algorithms as well as the time complexity bounds of $O(VE^2)$ and $O(V^2E)$, respectively. 
The formalization is conducted in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification to separate the abstract algorithmic ideas from
the implementation details: After proving the min-cut-max-flow theorem, it is straightforward to verify a generic version of the Ford-Fulkerson algorithm,
which we then refine to Edmonds-Karp algorithm and finally to an efficient implementation. 
Similarly, the generic push-relabel algorithm is refined to both, the relabel-to-front algorithm~\cite{CLRS09} and the FIFO push-relabel algorithm~\cite{GoTa88},
which, in turn, are refined to efficient implementations. The efficiency of our verified implementations compare well to unverified reference implementations of 
the algorithms.

The abstract parts of our verification closely follow the textbook presentation of Cormen et al.~\cite{CLRS09}. 
Using the Isar~\cite{Wenzel99} proof language, we were able to produce proofs that are accessible even to non-Isabelle experts.

This paper is an extended version of our conference paper~\cite{LaSe16}, which only covered the Edmonds-Karp algorithm.
While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial time maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation. Moreover, this paper is a case study on elegantly formalizing algorithms.

In the rest of this paper, we give a short informal introduction to flow networks and the Ford-Fulkerson method (\S\ref{sec:background}).
We then present our formalization of these material (\S\ref{sec:abs-formalization}). After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm, and on the complexity proof (\S\ref{sec:edka}).
Next, we describe the correctness and complexity proof of the generic push-relabel algorithm (\S\ref{sec:gen-prpu}), and its instantiations 
to the relabel-to-front and FIFO push-relabel algorithms (\S\ref{sec:prpu-inst}).
We then report on the refinement of all three algorithms to efficient implementations (\S\ref{sec:executable}), and their comparison to unverified reference 
implementations (\S\ref{sec:benchmark}).

The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}. 
The part on Edmonds-Karp algorithm is also available in the Archive of Formal Proofs~\cite{LaSe16_afp}.

@@

A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of incoming flows equals the 
sum of outgoing flows.
% 
% \emph{Non-generation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
% We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
% Intuitively, a \emph{preflow} must not send more flow over an edge than its capacity allows, and only the source node may generate flow. 
% The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
% A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all vertices except $s$ and $t$, the excess flow is zero.
% Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

@@

The correctness of the Ford-Fulkerson algorithm follows from the Ford-Fulkerson theorem, which is also known as min-cut-max-flow theorem.
It states that the following three propositions are equivalent:
\begin{enumerate}
\item $f$ is a maximum flow in a network $G$.
\item there is no augmenting path in the residual graph $G_f$.
\item there is a cut $C$ in $G$ such that the capacity of $C$ is equal to the value of $f$.
\end{enumerate}

@@

In a typical program development, one first designs an initial version \isai{m_0} of
the algorithm and its specification \isai{P,Q}, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions \isai{m$_i$} of the algorithm,
proving \isai{m$_i$ <= \<Down>R$_i$ m$_{i-1}$}. Using transitivity and composability of refinement, one 
gets \isai{P ==> m$_i$ <= \<Down>R$_i$...R$_1$ spec Q}, showing the correctness of the refined algorithm.
If no data refinement is performed, \isai{R$_i$} is set to the identity relation, in which case \isai{\<Down>R$_i$} becomes the identity function.

@@

\section{The Generic Push-Relabel Algorithm}\label{sec:gen-prpu}
  We present our formalization of the generic push relabel algorithm of Goldberg and Tarjan~\cite{GoTa88}. 
  Again, our formalization closely follows the textbook presentation of Cormen et al.~\cite{CLRS09}.

  Push-relabel algorithms are based on a generalization of flows called \emph{preflows}: 
  The conservation constraint is relaxed to allow more incoming flow than outgoing flow on a vertex.
  For a vertex $u\in V$, the \emph{excess} $x_f(u)$ is the difference between incoming and outgoing flow. 
  For a preflow, the excess of any node but the source must not be negative.
  A non-sink vertex with positive excess is called \emph{active}. Thus, a preflow without any active vertices is actually a flow.

  The push-relabel algorithm maintains a preflow $f$ and a height-labeling $l: V \to \mathbb {N}$ of the nodes, 
  with the invariant that for every edge $(u, v)$ of the residual graph we have that $l(u) \le l(v) + 1$. 
  Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
  By generalizing the labeling invariant to paths, we get that for each path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
  This implies that there cannot be a path from $s$ to $t$: The length of any (simple) path is less than $|V|$, however, a valid labeling 
  implies a length greater than or equal to $|V|$. Upon termination of the algorithm, there will be no active nodes. 
  Thus, the preflow is actually a flow, and as we have no $s$--$t$ paths, the Ford-Fulkerson theorem implies that it is maximal.

  The push relabel algorithm initializes the heights of all nodes but the source to zero. Moreover, the outgoing edges of the source are 
  set to carry maximum flow, all other edges get zero flow. Then, the algorithm repeats the following operation as long as they apply:
  \begin{description}
    \item[$\textrm{push}(u,v)$] for an edge $(u,v)$ of the residual graph with $l(u) = l(v)+1$ and $u$ active: Move as much flow as possible from $u$ to $v$.
    The maximum possible flow is constrained by the residual capacity of the edge, and the excess flow available at $u$.
    \item[$\textrm{relabel}(u)$] for an active node $u$ from which no push operation is possible: Increase the height of $u$ to the minimum value such that a push
      becomes possible.
  \end{description}
  Obviously, if none of the operations apply, all nodes are inactive and the resulting preflow is a maximum flow.
  The algorithm is generic in the sense that it leaves open the order in which the operations are applied to the different nodes and edges.

\subsection{Formalization of the Generic Algorithm}
For the complexity reasoning, we will count the different types of operations. Thus, it is convenient to formalize the algorithm as a labeled transition system
over flows and heights, the labels indicating the performed operation:
\begin{lstlisting}
  inductive_set pr_lts :: "((preflow\<times>labeling) \<times> op_ty \<times> (preflow\<times>labeling)) set" 
  where
    "push_pre f l e ==> ((f,l),PUSH,(push_eff f e,l)) \<in> pr_lts"
  | "relabel_pre f l u ==> ((f,l),RELABEL,(f,relabel_eff f l u)) \<in> pr_lts"
\end{lstlisting}  
Here, the \isai{inductive_set} construction defines the least relation that can be derived by the specified rules. 
A push operation can be implemented in time $O(1)$, (only the flow on a single edge has to be modified) and a relabel operation can be implemented in time $O(V)$.
(to find the new height, we have to consider the heights of all successor nodes in the residual graph) We define 
\isai{cost PUSH == 1 | cost RELABEL == card(V)} and prove the following theorem:
\begin{lstlisting}
theorem (in Network) generic_preflow_push_OV2E_and_correct:
  assumes "((f_0, l_0), p, (f, l)) \<in> pr_lts^*" 
  shows "(\<sum>x<-p. cost x) \<le> 38 * (card V)$^2$ * card E"
    and "(f,l)\<notin>Domain pr_lts --> isMaxFlow f"
\end{lstlisting}  
where \isai$(f_0,l_0)$ is the initial flow and height-labeling and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
Intuitively, the cost of any (partial) run \isai$p$ from the initial state to $(f,l)$ is bounded by $O(V^2E)$, and if the 
run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), the resulting flow is maximal.

For the proof of this theorem, we follow the presentation of Cormen et al.~\cite{CLRS09}.
The correctness proof shows that the algorithm maintains a valid labeling, and its formalization poses no further difficulties. 
It consists of less than 500 lines, and took us only a few days to complete.

The complexity argument is more tricky. Instead of defining a single measure function,
Cormen counts the different types of operations separately, using quite involved arguments.
For our formalization, we have adapted some Cormen's proofs, trading precision of the constant factors for simplicity of the formalization. 
Thus, our constant of $38$ is somewhat higher than the $17$ that would result from Cormen's proof.

First, we show that the labels of each node remain smaller than $2|V|$. This bounds the number of relabel operations to $O(V^2)$.
For the push operations, we define a push to be \emph{saturating}, if the excess at the source node of the push is not less than the capacity of the edge.
In this case, the edge is completely saturated by the push and disappears from the residual graph. However, the source node of the push remains active.
On the other hand, a non-saturating push deactivates the source node of the push.

We count the number of saturating and non-saturating pushes separately. For saturating pushes, we regard the number of saturating pushes for each edge $(u,v)$ during execution of the algorithm: After a saturating push, the edge has disappeared from the residual graph. Thus, before the next saturating push on the same edge, 
there must be an operation that restores the edge. This can only be a push on the reverse edge $(v,e)$. 
However, this push can only occur after the labeling of $v$ has increased. This can only happen $O(V)$ times, which yields a bound of $O(VE)$ on the number 
of saturating push operations. 
Finally, the number of non-saturating pushes is bounded by defining a potential function that is decreased by a non-saturating push.
A saturating push or relabel operation increases the potential by $O(V)$.
The initial potential is $O(V)$, and with the already established bounds on the other operations we get a bound of
$O(V^3 + V^2E) = O(V^2E)$ on the number of non-saturating push operations. Combining the bounds for the different operation types with the costs for the operations 
then yields the desired time complexity bound of $O(V^2E)$.

The simple structure of the generic algorithm allowed us to conveniently define it as labeled transition system, such that the above informal proof could be formalized
by inductive arguments over the runs of the algorithm.
Moreover, the powerful Isabelle/HOL standard library, in particular the lemmas for summations
over finite sets, were very helpful to keep the formalization effort manageable. 
Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few weeks to complete.
This includes the time that we spent on an initial termination proof that used simpler arguments, but did not derive an exact complexity bound.

\section{Instantiating the Generic Push-Relabel Algorithm}\label{sec:prpu-inst}
The generic algorithm does not specify the order in which the operations are applied. 
Choosing the right order decreases the worst-case time complexity, and, more important, the practical efficiency of the algorithm.

We consider two common instantiations of the generic algorithm: The relabel-to-front algorithm~\cite{CLRS09}, and tzhe FIFO push-relabel algorithm~\cite{GoTa88}.
Both have a theoretical time complexity of $O(V^3)$, which is better than the $O(V^2E)$ bound of the generic algorithm.
However, the more complicated structure of these algorithms prevent a convenient specification as labeled transition system, and thus 
a complexity proof in the style of the generic algorithm.
Instead, we use the Monadic Refinement Framework to specify the algorithms and only prove correctness and 
termination.\footnote{Which we get for free from the bound of the generic algorithm.} We leave it to future work to integrate complexity reasoning tools into 
the Refinement Framework and prove the tighter complexity bounds for the specific implementations.

A general idea of most efficient implementations is to \emph{discharge} a node by repeatedly applying push operations until the node gets inactive or needs to be relabeled.

\subsection{Relabel-to-Front Algorithm}
Cormen et al.\ present the relabel to front algorithm.
It discharges a node until it gets inactive, applying relabel operations as necessary. 
Moreover, if a node was discharged without being relabeled, the already saturated edges during this discharge are stored, and not considered 
during the next discharge. This reduces the work required to find the next edge to push.
The algorithm scans through a list of all nodes, discharging every node as it is visited.
If the node gets relabeled during discharging, it is moved to the front of the list, and the scan of the list begins anew.
The algorithm terminates if it reaches the end of the list.

@@

Following the proof by Cormen et al., and using the Isabelle Refinement Framework to conveniently describe the abstract algorithm,
the correctness proof offered no surprises. It is about 700 lines of Isabelle text plus 150 lines of general lemmas on topological orderings, 
and took only a few days to develop.

\subsection{FIFO Push-Relabel Algorithm}
While the relabel-to-front algorithm has a theoretical worst case complexity of $O(V^3)$, it seems to be rather inefficient in practice.\footnote{We have not found any implementation based on relabel-to-front, and our own experiments indicate a rather poor performance.}
An alternative implementation is the FIFO push-relabel algorithm, which maintains a first-in-first-out queue of the active nodes. 
In each iteration, it dequeues a node from the FIFO, and tries to discharge it by only using push operations.
If the node is still active, it is relabeled and enqueued.

The FIFO push-relabel algorithm has several advantages: It's correctness is straightforward to prove, 
as the only required additional invariant is that the queue contains exactly the active nodes. 
Moreover, it performs well in practice, and we have found several implementations of it~\cite{ChGo97,ICPC-notebook}.

Our algorithm follows the implementation that we found in the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, 
which uses an additional gap heuristics: If there is a value $0<k<|V|$ such that there is no node that has height $k$, 
all nodes with $k<l(v)<|V|$ may be relabeled to $|V|+1$, preserving a valid labeling. 
This heuristics prevents unnecessary pushes of flow towards the sink when the path to the sink is already blocked.
The correctness of the heuristics follows easily from the definitions of a valid height-labeling.
In total, the correctness proof of the FIFO push-relabel algorithm is about 400 lines of code and, building on the experience from relabel-to-front, it 
took only a day to complete.
